Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-06-30 workflow/workflow.Rmd
Updates to text.

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis alonglineages. Along the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.

@@

This workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium (OE) [@Fletcher2017]. The olfactory epithelium contains mature olfactory sensory neurons that are continuously renewed in the epithelium via neurogenesis through the differentiation of globose basal cells (GBCs), which are the actively proliferating cells in the epithelium. When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs) which become activated to differentiate and reconstitute all major cell types in the epithelium.

The scRNA-seq dataset we use as case study was generated to study the differentitation of HBC stem cells into different cell types present in the olfactory epithelium. To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing. Then, the expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it. Cells were then assigned to different lineages using a statistical analysis pipeline analogous to that in the present workflow. Finally, results were validated experimentally using in vivo lineage tracing. Details on data generation and statistical methods are available in [@Fletcher2017],  [@Risso2017], and [@Street2017].

It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs. Then, the GBC lineage, in turn, branches off to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman gland (Figure \@ref(fig:stemcelldiff)_. In this study, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.

@@

Throughout the workflow, we use an object of class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object. The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017]. Cells with a cluster label of `-2` were not assigned to a cluster in the original publication.

@@

Using the Bioconductor R package `scone`, we remove low-quality cells that did not pass our quality control filter, that is ... See Figure \@ref(fig:scone). Davide, would you like to describe your choices for scone here?

@@

Finally, to speed up the computations, we retain only the 1,000 most variable genes. We recommend, however, not to be so stringent in an actual data analysis.

@@

In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime ordering, and differential expression analysis. This allows the data to become more tractable, both from a statistical (cf. curse of dimensionality) and computational point of view. Additionally, technical noise can be reduced while preserving the often intrinsically low-dimensional signal of interest [@Peer2017] [@Risso2017][@Pierson2015].

Here, we perform dimensionality reduction using the Bioconductor R package `zinbwave`, which fits a zero-inflated negative binomial model (ZINB-WaVE) that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a sample-level intercept, which serves as a global-scaling normalization factor. The user can also include both gene-level and sample-level covariates. The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects, such as gene length and GC-content effects. A schematic view of the ZINB-WaVE model is provided in Figure \@ref(fig:zinbschema). For greated detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].

@@

The function `zinbDimRed` returns normalized expression measures as deviance residuals from the fit of the ZINB-WaVE model with user-supplied gene- and cell-level covariates. Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots). Note that in this case no low-dimensional matrix is computed, as some of the biological signal of interest could be removed when computing residuals with respect to this matrix.

@@

The principal component analysis (PCA) of the normalized values shows that, as expected, cells do not cluster by batch but by the orginal clusters (Figure \@ref(fig:pcanorm)). Overall, it seems that normalization was effective at removing batch effects.

@@

The `zinbDimRed` function can also be used to perform dimensionality reduction, where in this workflow the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`. The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance. To verify that `W` indeed captures the biological signal of interest, we display the MDS results in a scatterplot with colors corresponding to the original published clusters (Figure \@ref(fig:mdsW)).

>>>>

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis along lineages. Throughout the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.

@@

This workflow is illustrated using data from a scRNA-seq study of stem cell differentiation in the mouse olfactory epithelium (OE) [@Fletcher2017]. The olfactory epithelium contains mature olfactory sensory neurons that are continuously renewed in the epithelium via neurogenesis through the differentiation of globose basal cells (GBCs), which are the actively proliferating cells in the epithelium. When a severe injury to the entire tissue happens, the olfactory epithelium can regenerate from normally quiescent stem cells called horizontal basal cells (HBCs), which become activated to differentiate and reconstitute all major cell types in the epithelium.

The scRNA-seq dataset we use as a case study was generated to study the differentitation of HBC stem cells into different cell types present in the olfactory epithelium. To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing. The expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it. Cells were then assigned to different lineages using a statistical analysis pipeline analogous to that in the present workflow. Finally, results were validated experimentally using in vivo lineage tracing. Details on data generation and statistical methods are available in [@Fletcher2017],  [@Risso2017], and [@Street2017].

It was found that the first major bifurcation in the HBC lineage trajectory occurs prior to cell division, producing either mature sustentacular (mSUS) cells or GBCs. Then, the GBC lineage, in turn, branches off to give rise to mature olfactory sensory neurons (mOSN), microvillous (MV) cells, and cells of the Bowman gland (Figure \@ref(fig:stemcelldiff)). In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.

@@

Throughout the workflow, we use an object of class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object. The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017]. Cells with a cluster label of `-2` were not assigned to any cluster in the original publication.

@@

Using the Bioconductor R package `scone`, we remove low-quality cells that do not pass our quality control filter, that is ... See Figure \@ref(fig:scone). Davide, would you like to describe your choices for scone here?

@@

Finally, to speed up computation, we retain only the 1,000 most variable genes. We recommend, however, not to be so stringent in an actual data analysis.

@@

In scRNA-seq analysis, dimensionality reduction is often used as a preliminary step prior to downstream analyses, such as clustering, cell lineage and pseudotime ordering, and the identification of DE genes. This allows the data to become more tractable, both from a statistical (cf. curse of dimensionality) and computational point of view. Additionally, technical noise can be reduced while preserving the often intrinsically low-dimensional signal of interest [@Peer2017] [@Risso2017][@Pierson2015].

Here, we perform dimensionality reduction using the zero-inflated negative binomial-based wanted variation extraction (ZINB-WaVE) method implemented in the Bioconductor R package `zinbwave`. The method fits a ZINB model that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a sample-level intercept, which serves as a global-scaling normalization factor. The user can also include both gene-level and sample-level covariates. The inclusion of observed and unobserved sample-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects (e.g., gene length and GC-content effects). A schematic view of the ZINB-WaVE model is provided in Figure \@ref(fig:zinbschema). For greated detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].

@@

The function `zinbDimRed` returns normalized expression measures as deviance residuals from the fit of the ZINB-WaVE model, with user-supplied gene- and cell-level covariates. Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots). Note that in this case no low-dimensional matrix `W` is computed, as some of the biological signal of interest could be removed when computing residuals with respect to this matrix.

@@

The principal component analysis (PCA) of the normalized values shows that, as expected, cells do not cluster by batch but by the original clusters (Figure \@ref(fig:pcanorm)). Overall, it seems that normalization was effective at removing batch effects.

@@

The `zinbDimRed` function can also be used to perform dimensionality reduction, where, in this workflow, the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`. The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance. To verify that `W` indeed captures the biological signal of interest, we display the MDS results in a scatterplot with colors corresponding to the original published clusters (Figure \@ref(fig:mdsW)).
